### 2023-04-16 tex/article.tex
Up

Ocean fronts are narrow zones where horizontal gradients in water properties (temperature, salinity, nutrients, etc.) are significant \citep{belkin_2009a}, and are sometimes described as discontinuities because of their abrupt nature \citep{mauzole_2022}.
\Textcite{levy_2018} distinguished two types of fronts. Persistent fronts like those associated with the Gulf Stream and Kuroshio are locked in place by the coastal boundary and large-scale atmospheric forcing.
Their forcing is directly balanced by submesoscale symmetric instability, which takes energy mostly from the kinetic energy of the jet, and baroclinic instability, which converts the potential energy of the sloping density surfaces into large meanders and eddies.
Their ephemeral cousins are continuously forming, moving, and dissipating at the ocean surface.
They are being strained by mesoscale eddies, which intensifies their geostrophic along-front currents, and feeds fluid instabilities generating submesoscale vortices and filaments.
Importantly, a front will generate a cross-frontal ageostrophic secondary circulation \encadra*{an overturning circulation directed in the direction of trying to flatten the density surfaces in the front, with upwelling on the warm side of the front, and downwelling on the cold side (\cite{thomas_2008, mcwilliams_2016, mahadevan_2020})}.
The highly energetic persistent fronts are characterized by a deep vertical velocity structure that reaches the thermocline, while ephemeral fronts have associated cross-frontal secondary circulations that are generally confined to the vertically well-mixed upper layer of the ocean.
Frontal dynamics also involves increased stratification, reduction of mixed-layer depth and suppression of vertical mixing at the front (\cite{thomas_2008a}).

@@

In order to quantify how \as{chl} is affected by fronts at the local scale of fronts, we compared the distributions of \as{chl} over fronts and the distributions of \as{chl} outside of fronts.
Distributions were computed over 8~days windows to limit the influence of particularly cloudy days.
They were computed within each biome, and over latitudinal bands of width \ang{5}, to minimize the influence of the large-scale north south gradient in \as{chl}.
Distributions were compared in terms of their median value, but using mean values yielded similar results (see suppl.\ \nref{fig:hist-chl} for seasonal distributions and their medians over and outside of fronts within the three biomes).
We defined and computed the local excess \as{chl}, \as{excess} (expressed in per cent), as the median value over fronts minus the median value in the background, divided by the median value in the background, for each distribution.
We repeated this for weak and strong fronts.

To quantify the large-scale impact of fronts on \as{chl} at the scale of biomes, we computed the biome surplus of \as{chl} \as{surplus} which we defined and computed as the relative difference (expressed in per cent) between the mean \as{chl} over the entire biome (\(\mathit{MT}\)), and the mean \as{chl} over the background (\(\mathit{MB}\)), \(\am{surplus} = (\mathit{MT}-\mathit{MB}) / \mathit{MB}\).
Thus the surplus~\as{surplus} measures the extra quantity of \as{chl} at the scale of the biome (\(\mathit{MT} - \mathit{MB}\)), relative to what would be the situation in the absence of fronts (\(\mathit{MB}\)), and thus accounts for the local excess over fronts (\as{excess}), but also for the proportion of fronts in a given biome.
To better understand the meaning of the surplus, let us consider the simplified case where \as{chl} is homogeneously doubled over fronts compared to the background value, i.e.\ when \(\am{excess} = \pct{100}\); in that case, the surplus \as{chl} is \pct{50} if there are \pct{50} of fronts, and is \pct{1} if there are \pct{1} of fronts.
Note that the computation of the local excess~\as{excess} is based on median values because it relies on the comparison of distributions, while for the computation of the biome surplus~\as{surplus}, we used mean values in order to be conservative.

@@

Within each biome, the strength of the local excess in \as{chl} over fronts varies with latitude.
In the permanent subtropical biome (\nref{fig:latbands-s}), the excess is only detectable north of \latlon{25N} and remains modest (\(\am{excess} < \pct{10}\)); it reaches its maximum of \pct{10} at the beginning and end of the production season.
In the seasonal subtropical biome (\nref{fig:latbands-i}), the excess is also small in the southern part of the biome (south of \latlon{35N}) but strongly increases further north (between \latlon{35N} and \latlon{45N}) and decreases again going even further North (between \latlon{40N} and \latlon{45N}).
Moreover, in this biome, the increase is stronger in summer (July--August) compared to winter (December--January).
Finally in the subpolar biome, the magnitude of the local excess diminishes going northward.
It tends to be stronger during winter and during the bloom, and weaker in summer.
In fact in summer, the differences are hardly discernible when averaged over the entire biome (\nref{fig:ts-climato}c), but there is an excess \as{chl} in the southern part of the biome (\latlon{<45N}), and a small deficit in the northern part of the biome (\latlon{>45N}) (\nref{fig:latbands-n}).
It ensues that the annual mean local excess in \as{chl} over fronts has a distinct latitudinal pattern (\nref{fig:recap}a).
The strongest local increase due to fronts is located in the latitudinal band \latlonRange{35}{45N}, where the seasonal subtropical and subpolar biomes meet.
The excess is two to three times larger over fronts North of the Gulf Stream (i.e.\ in the subpolar biome) than South of it (i.e.\ in the seasonal subtropical biome).
Overall, the annual mean local excess over weak fronts varies between \num{0} and \pct{+30}, and between \pct{15} and \pct{+60} over strong fronts.

>>>>

Ocean fronts are narrow zones where horizontal gradients in water properties (temperature, salinity, nutrients, etc.) are significant (\cite{belkin_2009a}), and are sometimes described as discontinuities because of their abrupt nature (\cite{mauzole_2022}).
\Textcite{levy_2018} distinguished two types of fronts. Persistent fronts like those associated with the Gulf Stream and Kuroshio are locked in place by the coastal boundary and large-scale atmospheric forcing.
Their forcing is directly balanced by submesoscale symmetric instability, which takes energy mostly from the kinetic energy of the jet, and baroclinic instability, which converts the potential energy of the sloping density surfaces into large meanders and eddies.
Their ephemeral cousins are continuously forming, moving, and dissipating at the ocean surface.
They are being strained by mesoscale eddies, which intensifies their geostrophic along-front currents, and feeds fluid instabilities generating submesoscale vortices and filaments.
Importantly, a front will generate a cross-frontal ageostrophic secondary circulation \encadra*{an overturning circulation directed in the direction of trying to flatten the density surfaces in the front, with upwelling on the warm side of the front, and downwelling on the cold side (\cite{thomas_2008, mcwilliams_2016, mahadevan_2020})}.
The highly energetic persistent fronts are characterized by a deep vertical velocity structure that reaches the thermocline, while ephemeral fronts have associated cross-frontal secondary circulations that are generally confined to the vertically well-mixed upper layer of the ocean.
Frontal dynamics also involves increased stratification, reduction of mixed-layer depth and suppression of vertical mixing at the front (\cite{thomas_2008a}).

@@

In order to quantify how \as{chl} is affected by fronts at the local scale of fronts, we compared the distributions of \as{chl} over fronts and the distributions of \as{chl} outside of fronts.
Distributions were computed over 8~days windows to limit the influence of particularly cloudy days.
They were computed within each biome, and over latitudinal bands of width \ang{5}, to minimize the influence of the large-scale north south gradient in \as{chl}.
Distributions were compared in terms of their median value, but using mean values yielded similar results (see suppl.\ \nref{fig:hist-chl} for seasonal distributions and their medians over and outside of fronts within the three biomes).
We defined and computed the local excess \as{chl}, \as{exces} (expressed in per cent), as the median value over fronts minus the median value in the background, divided by the median value in the background, for each distribution.
We repeated this for weak and strong fronts.

To quantify the large-scale impact of fronts on \as{chl} at the scale of biomes, we computed the biome surplus of \as{chl} \as{surplus} which we defined and computed as the relative difference (expressed in per cent) between the mean \as{chl} over the entire biome (\(\mathit{MT}\)), and the mean \as{chl} over the background (\(\mathit{MB}\)), \(\am{surplus} = (\mathit{MT}-\mathit{MB}) / \mathit{MB}\).
Thus the surplus~\as{surplus} measures the extra quantity of \as{chl} at the scale of the biome (\(\mathit{MT} - \mathit{MB}\)), relative to what would be the situation in the absence of fronts (\(\mathit{MB}\)), and thus accounts for the local excess over fronts (\as{exces}), but also for the proportion of fronts in a given biome.
To better understand the meaning of the surplus, let us consider the simplified case where \as{chl} is homogeneously doubled over fronts compared to the background value, i.e.\ when \(\am{exces} = \pct{100}\); in that case, the surplus \as{chl} is \pct{50} if there are \pct{50} of fronts, and is \pct{1} if there are \pct{1} of fronts.
Note that the computation of the local excess~\as{exces} is based on median values because it relies on the comparison of distributions, while for the computation of the biome surplus~\as{surplus}, we used mean values in order to be conservative.

@@

Within each biome, the strength of the local excess in \as{chl} over fronts varies with latitude.
In the permanent subtropical biome (\nref{fig:latbands-s}), the excess is only detectable north of \latlon{25N} and remains modest (\(\am{exces} < \pct{10}\)); it reaches its maximum of \pct{10} at the beginning and end of the production season.
In the seasonal subtropical biome (\nref{fig:latbands-i}), the excess is also small in the southern part of the biome (south of \latlon{35N}) but strongly increases further north (between \latlon{35N} and \latlon{45N}) and decreases again going even further North (between \latlon{40N} and \latlon{45N}).
Moreover, in this biome, the increase is stronger in summer (July--August) compared to winter (December--January).
Finally in the subpolar biome, the magnitude of the local excess diminishes going northward.
It tends to be stronger during winter and during the bloom, and weaker in summer.
In fact in summer, the differences are hardly discernible when averaged over the entire biome (\nref{fig:ts-climato}c), but there is an excess \as{chl} in the southern part of the biome (\latlon{<45N}), and a small deficit in the northern part of the biome (\latlon{>45N}) (\nref{fig:latbands-n}).
It ensues that the annual mean local excess in \as{chl} over fronts has a distinct latitudinal pattern (\nref{fig:recap}a).
The strongest local increase due to fronts is located in the latitudinal band \latlonRange{35}{45N}, where the seasonal subtropical and subpolar biomes meet.
The excess is two to three times larger over fronts North of the Gulf Stream (i.e.\ in the subpolar biome) than South of it (i.e.\ in the seasonal subtropical biome).
Overall, the annual mean local excess over weak fronts varies between \num{0} and \pct{+30}, and between \pct{15} and \pct{+60} over strong fronts.

### 2023-04-16 tex/front_matter.tex
Add listings for CF xarray
Up

Force est de constater que, bien que les questions scientifiques posées dans ce manuscrit m'intéressent, m'entretenir à y répondre en tant que \guil*{chercheur} ne me correspond pas.
À l'inverse, j'ai pu trouver une réelle vocation dans le développement des outils requis par la résolution de ces questions.
Si vous me permettez cette analogie: si cette thèse n'aura pas pu me mettre sur les rails du jeune chercheur, elle m'a montré la voie de l'ingénieur de recherche, qui y court parallèle dans laquelle je me plais bien plus.
En conséquence, ce manuscrit reflète cette préférence et s'attardera peut"-être un peu plus qu'à l'accoutumée sur des considérations techniques.

@@

\begingroup
\defaultlists
Le travail effectué durant ma thèse et présenté dans ce manuscrit a donné lieu à plusieurs publications:
\begin{itemize}
  \setlength{\itemsep}{2\baselineskip}
  \renewcommand*\labelitemi{\adfrightarrowhead}
  \item un premier article présentant la méthode développée, et certains des résultats obtenus.
        Cet article est intégré à la structure de ce manuscrit, et inclus dans sa totalité au \nref[chapitre]{chp:res-chl}.
        \textnote{\fullcite{haeck_2023}}

  \item un second article décrivant l'étude effectuée par I.\ Mangolte dans la région du courant de Californie, et qui exploite en partie la méthode présentée dans cette thèse. Cet article est discuté dans le \hyperref[sec:CCE]{chapitre~\ref{chp:perspectives}} et resumé en \nref[annexe]{ax:article-cce}.
        \textnote{\review{\articleCceTitle}}

  \item et un article de \eng{review} dans lequel s’inscrit ce travail, discuté dans le \hyperref[sec:review]{chapitre~\ref{chp:perspectives}} et résumé en \nref[annexe]{ax:article-review}.
        \textnote{\review{\articleReviewTitle}}

@@

@@

Par ailleurs, pendant cette thèse une grande partie de mon effort s'est tournée sur l'élaboration d'outils informatiques cohérents \encadra{j'entends allant au"-delà d'une simple collection de scripts} permettant les analyses présentées ci"-après.
Une attention toute particulière a été portée à assurer leur ré"-utilisation et extension avec facilité, d'une part pour garantir la reproductibilité effective de cette étude, et d'autre part à ce que ce travail puisse bénéficier à d'autres.
Bien entendu, d'autres considérations entrent également en compte, par exemple de performances et de libre accès.
Tous les codes utilisés \encadra{accompagné qu'une documentation (partielle)} sont disponibles sur un dépôt Git public\footnote{\glsurl{gitlab}} hébergé par l'\abbrv{IN2P3}~(Institut National de Physique Nucléaire et de Physique des Particules, CNRS).
Une version a également été déposée sur un répertoire Zenodo (\hbox{\glsurl{zenodo}}).

@@

Certaines parties de ce travail en particulier se sont muées en projets indépendants, qu'il m'a paru utile de distribuer séparément.
J'ai ainsi durant ma thèse publié publiquement les outils suivant:
\begin{description}
  \setlength{\itemsep}{2\baselineskip}
  \item[FileFinder]
        est un paquet Python qui permet (entre autres) de trouver des fichiers grâce à la structure de leur nom de fichier, ce qui est particulièrement utile par exemple pour des fichiers dépendant de la date, ou de la valeur d'un paramètre, ou des deux.
        L'entièreté des fichiers de données utilisés pour cette thèse sont gérées grâce à cet outil.
        \textnote{\citesoftfullPerso{filefinder}}

  \item[tol-colors]
        est un paquet Python qui donne accès à des assortiments de couleurs adaptés aux personnes atteintes de daltonisme.
        Les jeux de couleurs étaient pré-existants et ont été développés par \glsname{paultol}\footnote{voir~\glsurl{paultol}}; je les ai seulement rendus plus facilement accessibles via PyPI.
        \textnote{\citesoftfullPerso{tol-colors}}

  \item[Xarray-histogram]
        est un paquet Python\footnote{%
        Il est à noter que ce projet est plus proche du stade de preuve de concept que d'outil exploitable.
        Son utilisation sur de large quantités de données avec Dask présente des problèmes non résolus.
        Néanmoins les bénéfices en performances sont très prometteurs (voir le dépôt Github pour des comparaisons).
        } qui permet de calculer des histogrammes de données gérées par \citesoft{xarray}.
        C'est une opération centrale dans les calculs effectués pour cette thèse (\nref{sec:extraction-hist}).
        Des solutions existent déjà, mais présentent certains désavantages (notamment de performances et de base de code complexe) que j'ai cherché à palier.
        \textnote{\citesoftfullPerso{xarray-histogram}}

  \item[dateloop]
        est une simple commande bash permettant de générer des ensembles de dates.
        Elle a surtout été utilisée au début de ce projet afin de faciliter la gestion et le téléchargement des fichiers de données par des scripts shell. Ces scripts ont progressivement été remplacés par des équivalents Python.
        \textnote{\citesoftfullPerso{dateloop}}

@@

De manière plus secondaire mais que je tiens néanmoins à noter, j'ai participé au projet \citesoft{cf-xarray} (visant à rendre accessible à \citesoft{xarray} les conventions de métadonnées \as{cf}) en y implémentant l'extraction de variables drapeaux sur masque binaire (\href{https://github.com/xarray-contrib/cf-xarray/pull/354}{\eng{pull request}~\textlf{\#354}}).
Le principe est détaillé en \nref[annexe]{ax:cf-flags}.

>>>>

Force est de constater que, bien que les questions scientifiques posées dans ce manuscrit m'intéressent, m'entretenir à y répondre en tant que \guil*{chercheur} ne me correspond pas.
À l'inverse, j'ai pu trouver une réelle vocation dans le développement des outils requis par la résolution de ces questions.
Si vous me permettez cette analogie: si cette thèse n'aura pas pu me mettre sur les rails du jeune chercheur, elle m'a montré la voie de l'ingénieur de recherche, qui y court parallèle dans laquelle je me plais bien plus.
En conséquence, ce manuscrit reflète cette préférence et s'attardera peut-être un peu plus qu'à l'accoutumée sur des considérations techniques.

@@

\begingroup
\defaultlists
Le travail effectué durant ma thèse et présenté dans ce manuscrit a donné lieu à plusieurs publications:
\begin{itemize}
  \setlength{\itemsep}{2\baselineskip}
  \renewcommand*\labelitemi{\adfrightarrowhead}
  \item un premier article présentant la méthode développée, et certains des résultats obtenus.
        Cet article est intégré à la structure de ce manuscrit, et inclus dans sa totalité au \nref[chapitre]{chp:res-chl}.
        \textnote[title={}]{\fullcite{haeck_2023}}

  \item un second article décrivant l'étude effectuée par I.\ Mangolte dans la région du courant de Californie, et qui exploite la méthode présentée dans cette thèse. Cet article est discuté dans le \hyperref[sec:CCE]{chapitre~\ref{chp:perspectives}} et resumé en \nref[annexe]{ax:article-cce}.
        \textnote[title={}]{\fullcite{mangolte_2023}}

  \item et un article de \eng{review} dans lequel s’inscrit ce travail, discuté dans le \hyperref[sec:review]{chapitre~\ref{chp:perspectives}} et résumé en \nref[annexe]{ax:article-review}.
        \textnote[title={}]{\review{\articleReviewTitle}}

@@

\review{Conferences}

@@

Par ailleurs, pendant cette thèse une grande partie de mon effort s'est tournée sur l'élaboration d'outils informatiques cohérents \encadra{j'entends allant au-delà d'une simple collection de scripts} permettant les analyses présentées ci-après.
Une attention toute particulière a été portée à assurer leur réutilisation et extension avec facilité, d'une part pour garantir la reproductibilité effective de cette étude, et d'autre part à ce que ce travail puisse bénéficier à d'autres.
Bien entendu, d'autres considérations entrent également en compte, par exemple de performances et de libre accès.
Tous les codes utilisés \encadra{accompagné d'une documentation} sont disponibles sur un dépôt Git public\footnote{\glsurl{gitlab}} hébergé par l'\abbrv{IN2P3}~(Institut National de Physique Nucléaire et de Physique des Particules, CNRS).
Une version a également été déposée sur un répertoire Zenodo (\hbox{\glsurl{zenodo}}).

@@

Certaines parties de ce travail en particulier se sont muées en projets indépendants, qu'il m'a paru utile de distribuer séparément.
J'ai ainsi durant ma thèse publié publiquement les outils suivant:
\begin{description}
  \setlength{\itemsep}{2\baselineskip}
  \item[FileFinder]
        est un paquet Python qui permet (entre autres) de trouver des fichiers grâce à la structure de leur nom de fichier, ce qui est particulièrement utile par exemple pour des fichiers dépendant de la date, ou de la valeur d'un paramètre, ou des deux.
        L'entièreté des fichiers de données utilisés pour cette thèse sont gérées grâce à cet outil.
        \textnote[title={}]{\citesoftfullPerso{filefinder}}

  \item[tol-colors]
        est un paquet Python qui donne accès à des assortiments de couleurs adaptés aux personnes atteintes de daltonisme.
        Les jeux de couleurs étaient pré-existants et ont été développés par \glsname{paultol}\footnote{voir~\glsurl{paultol}}; je les ai seulement rendus plus facilement accessibles via PyPI.
        \textnote[title={}]{\citesoftfullPerso{tol-colors}}

  \item[Xarray-histogram]
        est un paquet Python\footnote{%
        Il est à noter que ce projet est plus proche du stade de preuve de concept que d'outil exploitable.
        Son utilisation sur de large quantités de données avec Dask présente des problèmes non résolus.
        Néanmoins les bénéfices en performances sont très prometteurs (voir le dépôt Github pour des comparaisons).
        } qui permet de calculer des histogrammes de données gérées par \citesoft{xarray}.
        C'est une opération centrale dans les calculs effectués pour cette thèse (\nref{sec:extraction-hist}).
        Des solutions existent déjà, mais présentent certains désavantages (notamment de performances et de base de code complexe) que j'ai cherché à palier.
        \textnote[title={}]{\citesoftfullPerso{xarray-histogram}}

  \item[dateloop]
        est une simple commande bash permettant de générer des ensembles de dates.
        Elle a surtout été utilisée au début de ce projet afin de faciliter la gestion et le téléchargement des fichiers de données par des scripts shell. Ces scripts ont progressivement été remplacés par des équivalents Python.
        \textnote[title={}]{\citesoftfullPerso{dateloop}}

@@

De manière plus secondaire mais que je tiens néanmoins à noter, j'ai participé au projet \citesoft{cf-xarray} (visant à rendre accessible à \citesoft{xarray} les conventions de métadonnées \as{cf}) en y implémentant l'extraction de variables drapeaux sur masque binaire (\href{https://github.com/xarray-contrib/cf-xarray/pull/354}{\eng{pull request}~\textlf{\#354}}).
Le principe est détaillé en \nref[annexe]{ax:cf-flags}.
Cette technique sera d'ailleurs ré-utilisée pour extraire certains résultats (\nref{sec:extraction-hist}).
